Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-05-01 forecastflows.tex
integrate Lizzie edits and other stuff

\documentclass[11pt]{article}
\usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{multicol}
\usepackage{tcolorbox}
\usepackage{caption}
\usepackage{todonotes}
\usepackage[font=sf]{caption}

@@

\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology } 
\date{\today}
\author{Victor Van der Meersch, J. Regetz, T. J. Davies \& EM Wolkovich}
\maketitle

{\bf Deadline:} 1-ish May 2025 (was 1 April 2025, getting in by May 8th would be good!)

\emph{For:} Scientific and Statistical Workflow theme issue for \emph{Phil Trans A} as an \emph{Opinion}
% They mention a tex template here: https://royalsocietypublishing.org/rsta/for-authors#question4
% Word length: I should check what they emailed but they say no more than 13 printed pages with 650 per page (whoa, we should be well below that! I am thinking 3-5K would be good)
% Already 5K!

\begin{abstract}
Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. While new research often suggests that models accounting for these complexities can yield more accurate trends and predictions, we argue that the usual workflows for model fitting in ecology make it impossible to evaluate and compare the models used to generate insights and predictions. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could be transformative for ecological modeling.

@@

{\noindent \bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
\vspace*{0.5cm}

@@

Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on less data. Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from massive geographic, temporal and taxonomic biases. Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning. These approaches have failed to yield clear agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012}.

@@

Workflows that fully integrate all the steps required to build a model from an ecological question, with evaluation of limitations and potential problems before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge. Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

Quantitative science relies on a model-based framework to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and use this data to inform our model and differentiate between hypotheses. This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection. %add cites for prereg?
But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. In such cases, we must rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}. Indeed, most macroecological insights have emerged from exploring patterns in these datasets (exploratory data analysis).

\begin{figure}[h]
	\centering
	\includegraphics{figures/modeldataspaces_rotation}
	\caption{Different trajectories through model-data space during model development. Ideally, this begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is build---independently of the data. This is where we should spent most of the time, evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedbacks to refine our model.} 
	\label{fig:modeldata}
\end{figure}

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in term of data complexity and societal needs---instead may lead to persistent problems. % jdavies15april: what challenges? forecasting from imperfect data or the urgent need to generate forecasts?
Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Fig. \ref{fig:modeldata}a). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (Fig. \ref{fig:modeldata}b). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole. % jdavies15april: is this particularly specific to process based models? 
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.
% mention the disconnection between trends and forecasts?

% jdavies15april: Could the bottom right be where exploratory data analysis fits in, generating hypotheses that inform the research question?

We argue a workflow that moves along the data-model space in a coherent sequence of steps % with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure? Or wait to mention simulation until referencing Fig 2? [FLAG: what do you think? I agree you should choose on of the two options]
(Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, fig. \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error. % vvdm16april: to answer Jonathan comment about verbal model
From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, fig. \ref{fig:workflow}). The general idea is to start with a relatively simple model that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations).

With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, fig. \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data. % This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model. % jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e. simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model. [FLAG: yes I can understand this point...]
% For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. [FLAG: so I moved this sentence below, when we say "Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data"
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is reasonable or not. This is also a step for making sure the model is working as expected, so simulated data often would have low error and high sample sizes (to make sure the model returns the parameter values used to generate the simulated data). %jr24april not sure what this final sentence means?

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, fig. \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations. 
%emwApr19: I commented out the below as I think linking difficulty in fitting the model to problems with the data is pretty tricky (difficulty in fitting the model can point to many things!) and thus too confusing to include here. 
% Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question, or re-evaluate the model of how we think the system works. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, fig. \ref{fig:workflow}). This retrodictive check allows model output to be compared to observations. % jregetz16april: do we need to define “retrodictive”?
% The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting. % jregetz16april: Not sure what this means? jdavies did not like it either! (he thought it was a typo), vvdm16april: reformulate:
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting. 
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data. The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to more clearly develop a research question before extensive model fitting.

\begin{figure}[h]
	\centering
    \hspace*{-1.5cm}
	\includegraphics{figures/figure_worflow_wsteps}
	\caption{Here we could have a caption like ... Our workflow builds on others (with citations) to highlight the integrated nature of stuff in ecology. We can work on this once we get feedback from mighty co-authors. Also, should we list the steps on the figure? What do people think?} 
	\label{fig:workflow}
\end{figure}

% More details, what's special! Feedbacks! %jrApr25 I think this paragraph reads really well!
A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
flagging the need to change the model structure---before incorporating real observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es) (if we can identify them) and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

%emw6Apr: I like the shift to 'parameter estimation' instead of trend estimation, but I wonder if we could make that shift clearer above. Maybe somewhere we say that the workflow would  encourage a focus on the full model and its parameters: For trend estimates, the trend becomes just one parameter that must be carefully interpreted alongside others . 
Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to reduce uncertainties through new scientific insights, toward the most critical steps. This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data. Here, we illustrate how such a workflow could lead to significant improvements in two case studies: (i) estimating global biodiversity trends and (ii) forecasting future species and ecosystem dynamics using process-based models.

@@

%jregetz16april: One thing I’m stuck on (and could’ve written this note earlier too). Is “trends” really essential? Or is this just shorthand for “simple, one/few parameter, non-mechanistic forecasting models”? If yes, can we generalize that in the MS overall, or at least say so up front and then say e.g. “we’ll refer to these as trend models for brevity”? If no, and something about trend estimates is specifically key to the paper, can we say more about *that*? [FLAG]
%emwApr19: I think the only key part is that biodiversity trends and trends over time are big in forecasting ... as long as we don't lose that I think it would be great to make the suggested change -- Jim, can you see if you can edit it on the next round?

Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends over time, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}. While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced and more rapidly resolved through use of an improved workflow. % Further, an improved workflow could give more rapid and coherent estimates, which could make it easier for  policy-makers to develop and advance initiatives aimed at slowing declines.

@@

% expand on 'highlight major missing components in these models'? ADD example?
% Simulating data to verify and test models in trend analyses is currently rarely reported, which we believe makes it easier for ecologists to fit poor models and in turn debate them.. 
% Mention elephants rebounding...

% Trends - improvements!
Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates. Returning to the example of a global estimate of trends in vertebrate populations of species over time and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate? % jdavies15april: This is what the LPI wants to be (like a index fund in the stock market tracking trends over time) , but this is often overlooked.
Or a need to also find which species are declining most, including those that may have poor or no data? From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data. Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends. For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled. From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematical}.

@@

Ecological forecasting is a broad field with a diverse range of methods. Our second case-study focuses on process-based modeling, which is often considered the gold standard for forecasting in ecology \citep{Urban2016, Pilowsky2022} and beyond. Newer models generally incorporate greater complexity and an ever-growing number of parameters, making it more difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.

Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. \todo[size=\tiny]{jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting. Reply: we agree but not really relevant to PBM? Maybe it's not clear enough we are focusing on PBM in this section, which we tried to change above, see if it works}
If a model becomes too complex, it may become a black box, % mention black box here to line up with the beginning of next paragraph
and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}. \todo[size=\tiny]{jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty .... Reply: Same as just above.}

Process-based models used to project species or ecosystem distributions highlight some of these problems. Most of the focus is on the model projections---often represented as maps without uncertainties. Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only one of the many different output variables of the model. %The model building and calibration steps are usually limited to the Methods section, and estimated parameter values are rarely discussed \citep[e.g.][]{Conradi2024, VanderMeersch2025a}. 
The complexity of these models often makes it difficult to assess how well the potential parameter space was searched, and whether there was any potential non-identifiability---such work is generally done as a separate effort, exposing potential model problems later on \citep{VanderMeersch2025b}, and not required as a preliminary step. \todo[inline, size=\tiny]{For above, addressing: jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties? This could help illustrate potential degeneracies etc. Reply: Tried something, but not sure it works so feedback much appreciated!}

%emw19Apr -- Can we just get away with using non-identifiable and save the reader also needing to digest degenerate? (https://betanalpha.github.io/assets/case_studies/identifiability.html)? I tried that on this draft, see how it works!
Applying the workflow to process-based models would help open the black box.
Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two and help expose potential parameter identifiability issues in the model design. %jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below) 
Uncovering identifiability issues would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight model problems by requiring more explicit model calibration (i.e. data fitting), % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model. Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.

@@

% Trends - outline current problem

>>>>

\documentclass[11pt]{article}
\usepackage[top=0.70in, bottom=.80in, left=1in, right=1in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{multicol}
\usepackage{tcolorbox}
\usepackage{caption}
\usepackage{todonotes}
\usepackage[font=sf]{caption}
\usepackage{hyperref}

@@

\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology } 
\date{\today}
\author{Victor Van der Meersch$^{1*}$, James Regetz$^{2}$, T. Jonathan Davies$^{1,3}$ \& EM Wolkovich$^{1}$}
\maketitle
\noindent $^{1}$ Forest and Conservation Sciences, University of British Columbia, Vancouver, BC V6T 1Z4, Canada\\
$^{2}$ National Center for Ecological Analysis and Synthesis, 1021 Anacapa St, Santa Barbara, CA 93101, United States\\
$^{3}$ Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada \\
$^{*}$ \url{mailto:victor.vandermeersch@ubc.ca}\\

\noindent \emph{For:} Scientific and Statistical Workflow theme issue for \emph{Phil Trans A} as an \emph{Opinion}
% They mention a tex template here: https://royalsocietypublishing.org/rsta/for-authors#question4
% Word length: I should check what they emailed but they say no more than 13 printed pages with 650 per page (whoa, we should be well below that! I am thinking 3-5K would be good)
% Already 5K!

\begin{abstract}
Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.

@@

% {\noindent \bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.

@@

Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on less data. Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from large geographic, temporal and taxonomic biases. Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning. These approaches have failed to yield clear agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012, Thuiller2019}.

@@

Workflows that fully integrate all the steps required to build a model from an ecological question, with evaluation of limitations and potential problems before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge \citep{McElreath2018, betanworkflow, Gelman2020, Wolkovich2024}. Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

Quantitative science relies on a model-based framework to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and use this data to inform our model and differentiate between hypotheses. This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection \citep{Nosek2018}.
But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. In such cases, we must rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}. Indeed, most macroecological insights have emerged from exploring patterns in these datasets (exploratory data analysis).

\begin{figure}[h]
	\centering
	\includegraphics{figures/modeldataspaces_rotation}
	\caption{Different trajectories through model-data space during model development. Ideally, this begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is build---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedbacks to refine our model.}
	\label{fig:modeldata}
\end{figure}

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in term of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}a). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (figure \ref{fig:modeldata}b). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

We argue a workflow that moves along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error. 
From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}). The general idea is to start with a relatively simple model that we could refine later (see example worklow we provide). At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations).

With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data. 
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is reasonable or not. This is also a step for making sure the model is working as expected (see example worklow).

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations. 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}). This---which we call a retrodictive check (\citealp{betanworkflow}, also sometimes called a posterior predictive check, \citealp{Gelman2020})---allows model output to be compared to observations. The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.

Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data. The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to more clearly develop a research question before extensive model fitting.

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es) (if we can identify them) and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

\begin{figure}[h]
	\centering
	\vspace*{-0.1cm}
    \hspace*{-1.5cm}
	\includegraphics{figures/figure_worflow_wsteps}
	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data. Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5). Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.  The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.}
	\vspace*{-0.5cm}
	\label{fig:workflow}
\end{figure}

@@

Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to reduce uncertainties and yield new scientific insights. This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data. Here, we illustrate how such a workflow could lead to significant improvements in two case studies: (i) estimating global biodiversity trends and (ii) forecasting future species and ecosystem dynamics using process-based models.

@@

Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends over time, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}. While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be fewer and they would be more rapidly resolved through use of an improved workflow. % Further, an improved workflow could give more rapid and coherent estimates, which could make it easier for  policy-makers to develop and advance initiatives aimed at slowing declines.

@@

Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates. Returning to the example of a global estimate of trends in vertebrate populations of species over time (see Box) and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate?
Or a need to also find which species are declining most, including those that may have poor or no data? From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data. Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends (see workflow example). For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled. From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematical}.

@@

Ecological forecasting is a broad field with a diverse range of methods. Our second case-study focuses on process-based modeling, which is often considered the gold standard for forecasting in ecology \citep{Urban2016, Pilowsky2022} and beyond. Newer models, however, generally incorporate greater complexity (and an ever-growing number of parameters), which can make it difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.

Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for simplifying models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. 
If a model becomes too complex, it may become a black box,
and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

Process-based models used to project species or ecosystem distributions highlight some of these problems. Most of the focus is on the model projections---often represented as maps without uncertainties. Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only one of the many different output variables of the model.
The complexity of these models can make it difficult to assess how well the potential parameter space was searched, and whether there was any potential non-identifiability---such work is generally done as a separate effort, exposing potential model problems later on \citep{VanderMeersch2025b}, and not required as a preliminary step.

Applying the workflow to process-based models would help open the black box.
Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two, and help expose potential parameter identifiability issues in the model design. 
Uncovering identifiability issues would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight model problems by requiring more explicit model calibration (i.e. data fitting), which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model. Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.

@@

% Trends - outline current problem
In the current workflow for estimating trends over time, a new model with a new estimate often leads to a paper (see figure A above) because ecologists spend less time interrogating their models with simulated data, or their model performance fit to empirical data. 
The Living Planet Index (LPI, \url{www.livingplanetindex.org}), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results. % \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave. pop decline since 1970. This would be massive!}
With updated data released semi-annually alongside new datasets and new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For examples using LPI data, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed many previous results \citep{Johnson2024}.
